agdarsec -- Total Parser Combinators

### 2017-10-27 doc/paper/main.tex
[ cosmetic + references ]

\newcommand{\parser}[1]{\texttt{#1}}
\newcommand{\type}[1]{\texttt{#1}}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
\end{abstract}

@@

We start with a primer on parser combinators and follow up with the
definition of a broken parser which is silently accepted by Haskell.
We then move on to Agda~\todo{cite} and introduce combinators to
define functions by well-founded recursion. This allows us to define
a more informative notion of parser and give more precise types to
the combinators commonly used. We then demonstrate that broken parsers
such as the one presented earlier are rejected whilst typical example
can be ported with minimal modifications.

@@

\paragraph{Parser Type} Let us start by reminding ourselves what
a parser is. Although we will eventually move to a more generic
type, Fritz Ruehr's rhyme gives us the \emph{essence} of parsers:

@@

It is naturally possible to run such a parser and try to extract
a value from a successful parse. Opinions may vary on whether a
run with leftover characters can be considered successful or not.
We decide against it.\todo{ambiguity} This is not crucial as both styles can be
mutually emulated by either providing an 'end of file' parser
guaranteeing that only runs with no leftovers are successful or
by extending a grammar with a dummy case consuming the rest of
the input string.

@@

\paragraph{(Strongly-Typed) Combinators}

@@

However what makes parsers interesting is that they recognize
structure. As such, they need to reject invalid inputs. The
parser only accepting decimal digit is a bare bones example.
It can be implemented in terms of \parser{guard}, a higher
order parser checking that the value returned by its argument
abides by a predicate.

@@

These two definitions are only somewhat satisfactory: the
resut of the \parser{digit} parser is still stringly-typed.
Instead of using a predicate to decide whether to keep the
value, we can opt for a validation function which returns
a witness whenever the check succeeds. To define this
alternative \parser{guardM} we can once more rely on code
already part of the standard library.

@@

\paragraph{Code Reuse: Recognizing Structures}

We have seen how we can already rely on the standard library
of the host language to seemlessly implement combinators.
We can leverage even more of the existing codebase by noticing
that the type constructor \type{Parser} is a \type{Functor},
an \type{Applicative}~\todo{cite}, a \type{Monad} and also an \type{Alternative}.

@@

\paragraph{Expressivity: Higher Order Parsers and Fixpoints}

@@

@@

This approach can be generalised when defining more complex
languages by having even more tiers, one for each
\emph{fixity level}. An extended language of arithmetic
expressions would for instance distinguish the level at
which addition and substraction live from the one at which
multiplication and division do.

Our issue with this solution is twofold. Firt, although we did
eventually managed to build a parser that worked as expected,
the compiler was unable to warn us and guide us towards this
correct solution. Additionnally, the blatant partiality of
some of these definitions means that these combinators and
these types are wholly unsuitable in a total. We could, of
course use an escape hatch and implement our parsers in Haskell
but that would both be unsafe and mean we would not be able
to run them at typechecking time which we may want to do if
we embed checked examples in our software's documentation,
or use constant values (e.g. filepaths).

@@

The first kind of combinators corresponds to operations on
sets which are lifted to indexed sets by silently propagating
the index. We only show the ones we will use in this paper:
the pointwise arrow and product types and the constant function.
It may be a bit surprising to see infix notations (in Agda
underscores correspond to positions in which arguments are to
be inserted) for functions taking three arguments but they are
only meant to be partially applied.

@@

The second kind turns an indexed set into a set by universally
quantifying over the index. Here we use curly braces so that
the index we use is an \emph{implicit} argument we will never
have to write.

@@

@@

\remark{Counit} The careful reader will have noticed
that this is not quite the extract we would expect from a comonad:
for a counit, we would need a natural transformation between
\BOX{} \AB{A} and \AB{A} i.e. a function of type
\AF{[} \BOX{} \AB{A} \AF{⟶} \AB{A} \AF{]}. We will not be able
to define such a function: \BOX{} \AB{A} \AN{0} is isomorphic to
the unit type so we would have to generate an \AB{A} \AN{0} out
of thin air. The types \AB{A} for which \BOX{} has a counit are
interesting in their own right: they are inhabited at every single
index as demonstrated by \AF{fix} later on.

@@

\remark{Identifiers in Agda} Any space-free string which
is not a reserved keyword is a valid identifier. As a consequence
we can pick suggestive names such as \AB{m<n} for a proof that $m < n$.

@@

Exploring further the structure of the functor \BOX{}, we can observe
that just like it is not quite a comonad, it is not quite an applicative
functor. Indeed we can only define \AB{pure}, a natural transformation
of type \AF{[} \AB{A} \AF{⟶} \BOX{} \AB{A} \AF{]}, for the types \AB{A}
that are downwards closed. Providing the user with \AB{app} is however
possible:

@@

\remark{Generalisation} If we were to unfold all the type-level
combinators and record wrappers, the type of \AF{fix} would correspond
exactly to strong induction for the natural numbers. A similar \BOX{}
type constructor can be defined for any induction principle relying
on an accessibility predicate. Which means that a library's types
can be cleaned up by using these combinators in any situation where
one had to give up structural induction for a more powerful alternative.

@@

Unsurprisingly \AF{guardM} is still a valid higher-order combinator:
filtering out results which do not agree with a predicate is absolutely
compatible with the consumption constraint we have drawn. To implement
\AF{guardM} we can once more reuse existing library functions such
\AF{gfilter} which turns a \mbox{\AD{List} \AB{A}} into a \mbox{\AD{List} \AB{B}}
provided a predicate \mbox{\AB{A} \AS{→} \AD{Maybe} \AB{B}}.

@@

So far the types we have ascribed to our combinators are exactly the
same as the ones one would find in any other parsec library. In none
of the previous combinators do we run a second parser on the leftovers
of a first one. All we do is either manipulate or combine the results
of one or more parsers run in parallel, potentially discarding some of
these results on the way.

@@

The most basic example we can give of such an annotation is probably the
definition of a conjunction combinator \AF{\_<\&>\_} taking two parsers,
running sequentially and returning a pair of their results. The second
parser is given the type \mbox{\BOX{} \AR{Parser} \AB{B}} instead of
\mbox{\AR{Parser} \AB{B}} which we would expect to find in other parsec libraries.

@@

We can immediately use all of these newly-defined combinator to give
a safe, total definition of \AF{some} which takes a parser for \AB{A}
and returns a parser for non-empty lists of \AB{A}s. It is defined as
a fixpoint and proceeds as follows: it either combines a head and a
non-empty tail using \AF{$\_::^{+}\_$} or returns a singleton list.

@@

\remark{Inefficiency} Unfortunately this definition is inefficient.
Indeed, in the base case \mbox{\AF{some} \AB{p}} is going to run the parser
\AB{p} twice: once in the first branch before realising that \AB{rec}
fails and once again in the second branch. Compare this to the Haskell
version (after inlining \AB{many}):

@@

As the previous code snippet shows, \AF{\_<\&?>\_} is defined in
terms of a more fundamental notion \AF{\_\&?>>=\_} which is a
combinator analogous to a monad's \textit{bind}. On top of running
two parsers sequentially (with the second one being chosen based on
the result of running the first), it allows the second one to fail
and returns both results.

@@

The pattern used in the solution presented in section~\ref{sec:haskellproblem}
can be abstracted with the notion of an (heterogeneous) left chain which takes
a parser for a seed, one for a constructor, and one for and argument. The crucial
thing is to make sure \emph{not} to use the parser one is currently defining as
the seed.

@@

>>>>

\newcommand{\parser}[1]{\textit{#1}}
\newcommand{\type}[1]{\texttt{#1}}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Parser combinator libraries represent parsers as functions and,
using higher-order functions, define a DSL of combinators allowing
users to quickly put together programs capable of handling complex
recursive grammars. When moving to total functional languages such
as Agda, these programs cannot be directly ported: there is nothing
in the original definitions guaranteeing termination.

In this paper, we will introduce a `guarded' modal operator acting
on types and show how it allows us to give more precise types to
existing combinators thus guaranteeing totality. The resulting
library is available online together with various usage examples
at \url{https://github.com/gallais/agdarsec}.
\end{abstract}

@@

We start with a primer on parser combinators and follow up with the
definition of a broken parser which is silently accepted by Haskell.
We then move on to Agda~\cite{norell2009dependently} and introduce combinators to
define functions by well-founded recursion. This allows us to define
a more informative notion of parser and give more precise types to
the combinators commonly used. We then demonstrate that broken parsers
such as the one presented earlier are rejected whilst typical example
can be ported with minimal modifications.

@@

\subsection{Parser Type}

Let us start by reminding ourselves what a parser is. Although we
will eventually move to a more generic type, Fritz Ruehr's rhyme
gives us the \emph{essence} of parsers:

@@

It is naturally possible to run such a parser and try to extract
a value from a successful parse. Opinions may vary on whether a
run with leftover characters can be considered successful or not.
We decide against it. This is not crucial as both styles can be
mutually emulated by either providing an 'end of file' parser
guaranteeing that only runs with no leftovers are successful or
by extending a grammar with a dummy case consuming the rest of
the input string.

@@

\subsection{(Strongly-Typed) Combinators}

@@

However what makes parsers interesting is that they recognize
structure. As such, they need to reject invalid inputs. The
parser only accepting decimal digit is a bare bones example.
It can be implemented in terms of \parser{guard}, a higher
order parser checking that the value returned by its argument
abides by a predicate which can easily be implemented using
functions from the standard library.

@@

These two definitions are only somewhat satisfactory: the
resut of the \parser{digit} parser is still \emph{stringly-typed}.
Instead of using a predicate to decide whether to keep the
value, we can opt for a validation function of type
\mbox{\textit{a → Maybe b}} which returns
a witness whenever the check succeeds. To define this
alternative \parser{guardM} we can once more rely on code
already part of the standard library.

@@

\subsection{Code Reuse: Recognizing Structures}

We have seen how we can already rely on the standard library
of the host language to seemlessly implement combinators.
We can leverage even more of the existing codebase by noticing
that the type constructor \type{Parser} is a \type{Functor},
an \type{Applicative}~\cite{mcbride2008applicative}, a \type{Monad} and also an \type{Alternative}.

@@

\subsection{Expressivity: Higher Order Parsers and Fixpoints}

@@

\remark{Non-Commutativity} The disjunction combinator is
non-commutative. As such the definitions of \textit{some} and
\textit{many} will try to produce the longest list as possible
as opposed to a flipped version of \text{many} which would start
by returning the empty list and slowly offer longer and longer
matches.

@@

This approach can be generalised when defining more complex
languages by having even more tiers, one for each
\emph{fixity level}, see for instance~\ref{example:expr}.
An extended language of arithmetic expressions would for
instance distinguish the level at which addition and
substraction live from the one at which multiplication and
division do.

Our issue with this solution is twofold. Firt, although we did
eventually managed to build a parser that worked as expected,
the compiler was unable to warn us and guide us towards this
correct solution. Additionnally, the blatant partiality of
some of these definitions means that these combinators and
these types are wholly unsuitable in a total setting. We could, of
course use an escape hatch and implement our parsers in Haskell
but that would both be unsafe and mean we would not be able
to run them at typechecking time which we may want to do if
we embed checked examples in our software's documentation,
or use constant values (e.g. filepaths).

@@

The first kind of combinators corresponds to operations on
sets which are lifted to indexed sets by silently propagating
the index. We only show the ones we will use in this paper:
the pointwise arrow and product types and the constant function.

@@

\remark{Mixfix Operators} In Agda underscores correspond to
positions in which arguments are to be inserted. It may be a
bit surprising to see infix notations for functions taking
three arguments but they are only meant to be partially applied.
\medskip{}

The second kind of combinator turns an indexed set into a set
by universally quantifying over the index.

@@

\remark{Implicit Arguments} We use curly braces so that the
index we use is an \emph{implicit} argument we will never
have to write: Agda will fill it in for us by unification.
\medskip{}

@@

\remark{Counit} The careful reader will have noticed
that this is not quite the \textit{extract} we would expect from a
comonad: for a counit, we would need a natural transformation
between \BOX{} \AB{A} and \AB{A} i.e. a function of type
\mbox{\AF{[} \BOX{} \AB{A} \AF{⟶} \AB{A} \AF{]}}. We will not be able
to define such a function: \BOX{} \AB{A} \AN{0} is isomorphic to
the unit type so we would have to generate an \AB{A} \AN{0} out
of thin air. The types \AB{A} for which \BOX{} has a counit are
interesting in their own right: they are inhabited at every single
index as demonstrated by \AF{fix} later on.

@@

\remark{Identifiers in Agda} Any space-free string which
is not a reserved keyword is a valid identifier. As a consequence
we can pick suggestive names such as \AB{m<n} for a proof that $m < n$
(notice the extra spaces around the infix operator $(<)$).

@@

Exploring further the structure of the functor \BOX{}, we can observe
that just like it is not quite a comonad, it is not quite an applicative
functor. Indeed we can only define \AB{pure}, a natural transformation
of type \mbox{\AF{[} \AB{A} \AF{⟶} \AR{\BOX{}} \AB{A} \AF{]}}, for the types \AB{A}
that are downwards closed. Providing the user with \AB{app} is however
possible:

@@

If we were to unfold all the type-level combinators and record wrappers,
the type of \AF{fix} would correspond exactly to strong induction for
the natural numbers. Hence its implementation also follows the one of strong
induction: it is a combination of a call to \AF{extract} and an auxiliary
definition \AF{fix\BOX{}} of type \mbox{\AF{[} \AR{\BOX{}} \AB{A} \AF{⟶}
\AB{A} \AF{]} \AS{→} \AF{[} \AR{\BOX} \AB{A} \AF{]}}.

\remark{Generalisation}  A similar \BOX{}
type constructor can be defined for any induction principle relying
on an accessibility predicate. Which means that a library's types
can be cleaned up by using these combinators in any situation where
one had to give up structural induction for a more powerful alternative.

@@

Unsurprisingly \AF{guardM} is still a valid higher-order combinator:
filtering out results which do not agree with a predicate is absolutely
compatible with the consumption constraint we have drawn. To implement
\AF{guardM} we can once more reuse existing library functions such as
\AF{gfilter} which turns a \mbox{\AD{List} \AB{A}} into a \mbox{\AD{List} \AB{B}}
provided a predicate \mbox{\AB{A} \AS{→} \AD{Maybe} \AB{B}}.

@@

So far the types we have ascribed to our combinators are, if we ignore
the $\mathbb{N}$ indices, exactly the
same as the ones one would find in any other parsec library. In none
of the previous combinators do we run a second parser on the leftovers
of a first one. All we do is either manipulate or combine the results
of one or more parsers run in parallel, potentially discarding some of
these results on the way.

@@

The most basic example we can give of such an annotation is probably the
definition of a conjunction combinator \AF{\_<\&>\_} taking two parsers,
running them sequentially and returning a pair of their results. The second
parser is given the type \mbox{\BOX{} \AR{Parser} \AB{B}} instead of
\mbox{\AR{Parser} \AB{B}} which we would expect to find in other parsec libraries.

@@

We can immediately use all of these newly-defined combinators to give
a safe, total definition of \AF{some} which takes a parser for \AB{A}
and returns a parser for non-empty lists of \AB{A}s. It is defined as
a fixpoint and proceeds as follows: it either combines a head and a
non-empty tail using \AF{$\_::^{+}\_$} or returns a singleton list.

@@

\remark{Inefficiency} Unfortunately this definition is inefficient.
Indeed, in the base case \mbox{\AF{some} \AB{p}} is going to run the parser
\AB{p} twice: once in the first branch before realising that \AB{rec}
fails and once again in the second branch. Compare this definition to
the Haskell version (after inlining \AB{many}) where \AB{p} is run once
and then its result is either combined with a list obtained by recursion
or returned as a singleton:

@@

As the previous code snippet shows, \AF{\_<\&?>\_} is defined in
terms of a more fundamental notion \AF{\_\&?>>=\_} which is a
combinator analogous to a monad's \textit{bind}. On top of running
two parsers sequentially (with the second one being chosen based on
the result obtained by running the first), it allows the second one to fail
and returns both results.

@@

The pattern used in the solution presented in Section~\ref{sec:haskellproblem}
can be abstracted with the notion of an (heterogeneous) left chain which takes
a parser for a seed, one for a constructor, and one for and argument. The crucial
thing is to make sure \emph{not} to use the parser one is currently defining as
the seed.

@@

\subsection{Examples}

\todo{expr}\label{example:expr}
